Higher-Order unification for free!

### 2024-04-17 paper/main.tex
many todo to simplify

\noindent
Note that we could remove the first rule, whose sole purpose is to make
links more readable by pruning unused context entries.

@@

@@

\newcommand{\progressetashrink}{\emph{$\eta$-progress-shrink}\xspace}
\begin{definition}[\progressetashrink]\label{def:progressetashrink}
Given a link \linketaM{\Gamma}{X}{t}, and let $S$ the scope of $X$
we pose $T = S \cap \Gamma$. If $T$ is not empty,
then the original link is replaced by a new link
\linketaM{S}{X}{t'}, where $t'$ is a term where all the variables in $\Gamma / X$ 
have been pruned.
\end{definition}

@@

\begin{proof}[Proof sketch]\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  succeeds. A $\hstep_i$ is made by a unification $u$ and a link
  progression $p$.
  Without loss of generality,\todo{explain why} we consider a \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$, which is compiled into the \linketa $\linketaM{}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to treat the progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$, in this case, the unification problem becomes
          $X \Ue t_2$, which assigns $X$ to $t_2$.
          After the unification, \progressetaleft triggers the \linketa,
          and $\lambda x.t_1$ is unified with $t_2$ (or its $\eta$-expansion when needed).
          This unification succeeds iff $s_1$ and $s_2$ unify. \progressetaleft
          aims essentially to transpose the two rules \ref{clause:eta1} and \ref{clause:eta2}
          of the object language in the meta one.
    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and $t_1$ and $t_2$ are unified.
          This unification succeeds iff $s_1 \Uo s_2$ succeeds.
  \end{itemize}
  % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  % we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  % In all cases $X \Ue t2$ succeeds since X is a variable and
  % hence assigns t2 to X.
  % Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  % additional care of eta expanding t2 if needed (5.9.1).
% 
  % Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  % $s1$ with $s2$. The interesting case is when fostep succeds by
  % finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  % rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  % and that is in exactly the cases in which T (from page 3) is
  % not a lam (see 5.9.1) compensating for the missing rules in \Ue.
  % 
  % 2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  % and the one flam-flam we find different terms.
  % (5.9.1) only adds lambdas that are forcibly consumed by
  % lam-lam rule.
\end{proof}

@@

\paragraph{Example of \progressetadedup}

@@

\noindent
The result of $A \Ue{} C$ is that the two \linketa share the same \lhs.
By unifying the two \rhs we get
$\sigma = \{ A~ \mapsto C, B \mapsto D ~\}$.
In turn, given the map \mapStore, this second assignment is decompiled to
$\rho = \{~ X \mapsto Y ~\}$ as expected.

@@

In \cref{sec:grounwork}, we introduced the definition of memory map \mapStore.
This map allows to decompile the \Ho substitution to the object language.
Given the higher order nature of Coq the same unification
variable $X$ is used multiple times with different arities.
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

In the unification problems \foUnifPb above, we see that $X$ is used with arity $2$ in
$\foUnifPb_1$ and with arity $1$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, and in turn this
breaks \fullref{prop:fidelity} since two incompatible assignements to
the variables $X$ maps to, namely $E$ and $C$, would not be detected.
We remedy by post-processing the compiler's output by adding
additional \linketa to ensure the folllowgin property:

@@

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{t} is removed from \linkStore when either 1)
$\testmaybeeta~t$ does not hold (anymore) or 2) by $\eta$-contracting $t$ to
$t'$ where $t'$ does not start with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $t$, and in the second one, $X$ is unified with
$t'$ (under the context $\Gamma$).
\end{definition}

@@

\paragraph{Example of \progBetaLL}
Consider the \linkbeta below:
%
% make test ONLY=7004 TEX=tex
\printAlll
  {{{X,\lambda x.x},
    {\lambda x.(Y\appsep (X\appsep x)),f}}}
  {{{A,\lambda x.x},
    {B,f}}}
  {{{Y,D,0},
    {X,A,0}}}
  {{{\eta,,A,\lambda x.E_{x}},
    {\eta,,B,\lambda x.C_{x}},
    {\llambda,x,C_{x},(D\appsep E_{x})}}}

@@

The first unification problem is solved with the substitution
$\sigma = \{~A \mapsto \lambda x.a~\}$. The \linkbeta becomes
\linkbetaM{}{B}{((\lambda x.a) \appsep a)} whose \rhs can be $\beta$-reduced to
$a$. $a$ is in \llambda and is unified with $B$. The resolution of the second
unification problem
confirm that $B$ equal $a$. The final substitution $\sigma = \{~A \mapsto \lambda x.a ~;~ B \mapsto a ~\}$ is decompiled into $\rho = \{~ X \mapsto \lambda x.a ~;~ Y \mapsto
a ~\}$.

\subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}
\label{sec:heuristics}
\newcommand{\progBetaNoLLWait}{\emph{progress-beta-\notllambda}}

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R HolesInR, !,
  declare_constraint (link-eta L R) [L|HolesInR].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
\end{elpicode}
\todo{improve code}

>>>>

\noindent
\todo{remove pruning optimization from code above}

@@

\subsubsection{Scope check}\label{sec:sc}
The predicate xxxx \todo{add} takes
a link \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} and replaces it by
\linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ after
executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$, that is $t'$ cannot contain
variables in $\Gamma / \{x_1\ldots x_n\}$.
This is key for fidelity since suspended unif problem
must stay solvable.

@@

@@

\begin{proof}[Proof sketch]\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  succeeds. A $\hstep_i$ is made by a unification $u$ and a link
  progression $p$.
  Without loss of generality,\todo{explain why} we consider a
  \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$. $s_1$ is compiled to $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compiler to $t_2$ and $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft if $t_2$ is rigid.
          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they unify thanks to lam-lam\todo{dare nome}
          and $p$ unifies $\lambda x.t_1$ with $t_2$. Alternatively, if $s_1$ and $s_2$ fails
          to unify, it cannot be because of the head $\lambda$ constructors,
          since wither lam-lam or \ref{clause:eta1} move under it and find
          different sub terms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.

    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and, once the $\lambda$ are crossed,
          $t_1$ and $t_2$ are unified. 
          This unification succeeds iff $s_1 \Uo s_2$ succeeds by \cref{th:fidelity-beta}.
  \end{itemize}
  % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  % we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  % In all cases $X \Ue t2$ succeeds since X is a variable and
  % hence assigns t2 to X.
  % Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  % additional care of eta expanding t2 if needed (5.9.1).
% 
  % Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  % $s1$ with $s2$. The interesting case is when fostep succeds by
  % finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  % rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  % and that is in exactly the cases in which T (from page 3) is
  % not a lam (see 5.9.1) compensating for the missing rules in \Ue.
  % 
  % 2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  % and the one flam-flam we find different terms.
  % (5.9.1) only adds lambdas that are forcibly consumed by
  % lam-lam rule.
\end{proof}

@@

\begin{comment}
\paragraph{Example of \progressetadedup}

@@

\noindent
The result of $A \Ue{} C$ is that the two \linketa share the same \lhs.
By unifying the two \rhs we get
$\sigma = \{ A~ \mapsto C, B \mapsto D ~\}$.
In turn, given the map \mapStore, this second assignment is decompiled to
$\rho = \{~ X \mapsto Y ~\}$ as expected.
\end{comment}

@@

In \cref{sec:grounwork}, we introduced the definition of memory map \mapStore.
This map allows to decompile the \Ho substitution to the object language.
Given the higher order nature of Coq the same unification
variable $X$ is used multiple times with different arities.
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\todo{simplify example since no prog right}
In the unification problems \foUnifPb above, we see that $X$ is used with arity $2$ in
$\foUnifPb_1$ and with arity $1$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, and in turn this
breaks \fullref{prop:fidelity} since two incompatible assignements to
the variables $X$ maps to, namely $E$ and $C$, would not be detected.
We remedy by post-processing the compiler's output by adding
additional \linketa to ensure the folllowgin property:

@@

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{t} is removed from \linkStore when either 1)
$\testmaybeeta~t$ does not hold (anymore) or 2) by $\eta$-contracting $t$ to
$t'$ where $t'$ does not start with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $t$, and in the second one, $X$ is unified with
$t'$ (under the context $\Gamma$).
\end{definition}
\todo{comment this rule}

@@

\begin{comment}
\paragraph{Example of \progBetaLL}
Consider the \linkbeta below:
%
% make test ONLY=7004 TEX=tex
\printAlll
  {{{X,\lambda x.x},
    {\lambda x.(Y\appsep (X\appsep x)),f}}}
  {{{A,\lambda x.x},
    {B,f}}}
  {{{Y,D,0},
    {X,A,0}}}
  {{{\eta,,A,\lambda x.E_{x}},
    {\eta,,B,\lambda x.C_{x}},
    {\llambda,x,C_{x},(D\appsep E_{x})}}}

@@

The first unification problem is solved with the substitution
$\sigma = \{~A \mapsto \lambda x.a~\}$. The \linkbeta becomes
\linkbetaM{}{B}{((\lambda x.a) \appsep a)} whose \rhs can be $\beta$-reduced to
$a$. $a$ is in \llambda and is unified with $B$. The resolution of the second
unification problem
confirm that $B$ equal $a$. The final substitution $\sigma = \{~A \mapsto \lambda x.a ~;~ B \mapsto a ~\}$ is decompiled into $\rho = \{~ X \mapsto \lambda x.a ~;~ Y \mapsto
a ~\}$.
\end{comment} 
\subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}
\label{sec:heuristics}
\newcommand{\progBetaNoLLWait}{\emph{progress-beta-\notllambda}}

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R HolesInR, !,
  declare_constraint (link-eta L R) [L|HolesInR].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
\end{elpicode}
\todo{improve code}
%link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
